How many speech sounds are there in the world's languages?

### 2023-08-04 paper.tex
submitted version

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{22 June, 2023}

@@

How many speech sounds are there in the world's languages? To answer
this question, we extend two formal models of lexical diversity to
segmental phonology. Zipf's law states that most word types are rare;
Herdan-Heaps' law states that vocabulary size grows indefinitely with
increasingly new text. We find that both models produce accurate
estimates of the diversity of contrastive speech sounds (phonemes) in
existing typological databases. Our findings suggest that (a) most
phonemes are difficult to detect because they are rare, and (b)
documenting more phonological repertoires cross-linguistically always
yields more phoneme types. As a sanity check, we apply an unrelated
estimator of species diversity from biology, Chao-Shen, and show that
its estimates align with Herdan-Heaps' law. We use both models to
extrapolate phoneme counts to undocumented languages, resulting in an
estimate of 4600-4900 distinct phonemes in the full set of languages
thought to be spoken today (\textasciitilde7-8k), which is roughly 1.5x
more than what is currently documented in the world's languages
(\textasciitilde35\%). By extrapolating to hypothetical sample sizes,
e.g., the number of languages estimated to have been spoken through
time, and beyond, we find no indication of convergence: infinite
languages mean infinite phonemes.

@@

Assuming a target of 7151 known languages,\footnote{\url{https://www.ethnologue.com/guides/how-many-languages}?}
our models predict a range of total segment types from 4592 to 4850
(mean = 4721). Based on PHOIBLE, our current best guess of the diversity
of documented phoneme types is 3183. Therefore, the models argue for the
existence of approximately 1500 yet to be documented phoneme types. As
we probe increasingly larger and theoretically motivated language sample
sizes, i.e., languages that existed \emph{through time} (e.g., estimates
including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
Herdan-Heaps' model predicted a continual increase in the number of
unique phonemes, reaching values as high as 19K. The Chao-Shen model
produced much more conservative estimates for samples of size 10K and
above, even appearing to settle on an asymptote of roughly 5500 phonemes
for samples of 100K languages or more. However, when we compare the
predictions of Herdan-Heaps and Chao-Shen against known type counts in a
closed dataset, the source of the discrepancy emerged: as the sample
grows, Chao-Shen lags behind Herdan-Heaps in approximating the true
value. Thus, given a larger pool of training data, we expect Chao-Shen
to approximate more closely the estimates of Herdan-Heaps' law.

We therefore provisionally conclude three things: (1) adding languages
to a sample will always carry a non-zero probability of producing a
novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at
which new phonemes are discovered per language will always decrease; and
(3) the decrease in the rate of new observations will never reach 0.
Thus, based on the evidence presented here, we have no reason to believe
that there is an upper bound on the number of unique phonemes possible
in spoken languages. Infinite linguistic diversity seems to breed
infinite possible phonemes. But what would it mean if there were indeed
infinitely many potential phonemes?

If one adopts a feature-based perspective on the phonetic targets of
phonemes, then there is an \emph{a priori} upper bound of \(2^F\)
contrastive sounds, where \(F\) is the number of feature categories and
2 reflects the choice between presence (+) or absence (-) of that
feature.\footnote{In some feature sets, some values may be 0, i.e.,
  undefined, underspecified, or ``for which a segment can be said `not
  to care''' (\protect\hyperlink{ref-Hayes2009}{Hayes 2009}, pg. 91).
  For example, speech sounds that are -coronal are underspecified for
  the features +/-anterior and +/-distributed, which capture coronal
  distinctions for places of articulation involving (inter-)dentals and
  alveolars vs palatal alveolars and retroflexes (anterior), and laminal
  vs apical coronals (distributed).} Clements
(\protect\hyperlink{ref-Clements2009}{2009}) coined this calculation
\emph{Feature Bounding}. The expected number of unique phonemes is
controlled by the number of features, where additional features increase
the pool of possible distinctive segments (nearly) exponentially. For
example, given a feature set such as in PHOIBLE with 37 distinctive
features, an upper bound on the number of segments under a strictly
binary formalization is over 137 billion (2\^{}37) feature vectors.

However, most widely accepted feature sets contain features that are
specified at different levels of phonetic detail: some features are more
abstract (i.e., general similarities such as +/- syllabic), while others
are more concrete (i.e., closer to the articulatory reality, e.g.,
advanced tongue root). Some are grounded in pure articulatory or
acoustic phonetics, while others involve phonological properties of
language. And looming in the background of these discussions, the debate
of whether distinctive features are innate and universal or rather
learned and emergent continues. Nevertheless, in general, distinctive
feature theories of different theoretical stances propose around two
dozen features (\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume
2006}), which still has the potential to generate over 16 million
distinct feature vectors (2\^{}24). Current consensus aside, there does
not appear to be a principled limit to the degree of phonetic
specificity that can be encoded by a feature. Making finer and finer
contrasts may always produce more and more features. Thus, the exponent
in the Feature Bounding calculation can itself (by hypothesis) be
infinite, in line with the predictions of the models presented here, and
therefore with the notion of infinite possible phonemes. That said,
contemporary feature sets can capture phonological classes for place and
manner of articulation, which neatly classify the phonological patterns
and processes in and across languages (cf. Mielke
(\protect\hyperlink{ref-Mielke2008}{2008})).

We have so far seen that the feature space is a discrete re-coding of
the continuous variability of the physical acoustic and articulatory
spaces -- and so it should by way of finer and finer specification, be
able to support an infinitely diverse set of phonemes. However, one
objection to this account is that humans are subject to physiological
limits, both in production and perception. In speech production,
articulatory gestures are executed with some degree of imprecision
(\protect\hyperlink{ref-Ladefoged_etal1972}{Ladefoged et al. 1972};
\protect\hyperlink{ref-Lindblom1990}{Lindblom 1990}). If this stochastic
variability is greater than the difference between two or more phonetic
targets, then no reliable distinction can be sustained (cf. Stevens
(\protect\hyperlink{ref-Stevens1972}{1972})). In comparison, if the
range of articulatory gestures is greater than their acoustic (or
perceptual) reception, there is a ``many-to-one'' mapping between
articulation and perception. In perception, the ability to discriminate
acoustic stimuli is also limited both anatomically and cognitively.
Differences in the signal that fall below the various thresholds for
discrimination are thus ignored.

@@

Given that all speakers use perceptually contrastive cues (whether
signed, spoken or written), how these continuous signals map to discrete
elements is a core question of language science that goes beyond
whatever the phonetics-phonology interface may or may not be. Thus, we
believe the approach presented here has broad applications to the study
of other areas of grammar and to the cross-linguistic study of
languages.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BaezaNavarro2000}{}}%
Baeza--Yates, Ricardo \& Gonzalo Navarro. 2000. Block addressing indices
for approximate text retrieval. \emph{Journal of the American Society
for Information Science}. Wiley Online Library 51(1). 69--82.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ChaoChiu2016}{}}%
Chao, Anne \& Chun-Huo Chiu. 2016. Species richness: Estimation and
comparison. In \emph{Wiley StatsRef: Statistics reference online},
1--26. John Wiley \& Son.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Egghe2007}{}}%
Egghe, Leo. 2007. Untangling herdan's law and heaps' law: Mathematical
and informetric arguments. \emph{Journal of the American Society for
Information Science and Technology}. Wiley Online Library 58(5).
702--709.

\leavevmode\vadjust pre{\hypertarget{ref-EvansLevinson2009}{}}%
Evans, Nicholas \& Stephen C. Levinson. 2009. The myth of language
universals: Language diversity and its importance for cognitive science.
\emph{Behavioral and Brain Sciences}. Cambridge University Press 32(5).
429--448.

\leavevmode\vadjust pre{\hypertarget{ref-Everett2018}{}}%
Everett, Caleb. 2018. The similar rates of occurrence of consonants
across the world's languages: A quantitative analysis of phonetically
transcribed word lists. \emph{Language Sciences}. Elsevier 69. 125--135.

\leavevmode\vadjust pre{\hypertarget{ref-FirchowFirchow1969}{}}%
Firchow, Irwin \& Jacqueline Firchow. 1969. {An Abbreviated Phoneme
Inventory}. \emph{Anthropological Linguistics} 11. 271--276.

\leavevmode\vadjust pre{\hypertarget{ref-GoodCysouw2013}{}}%
Good, Jeff \& Michael Cysouw. 2013. Languoid, doculect, and glossonym:
Formalizing the notion'language'. \emph{Language Documentation and
Conservation} 7. 331--359.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Herdan1964}{}}%
Herdan, Gustav. 1964. Quantitative linguistics. Butterworth.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Joos1948}{}}%
Joos, Martin. 1948. Acoustic phonetics. \emph{Language}. Linguistic
Society of America 24(2). 5--136. (7 December, 2022).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Ladefoged_etal1972}{}}%
Ladefoged, Peter, Joseph DeClerk, Mona Lindau \& George Papcun. 1972. An
auditory-motor theory of speech production. \emph{{UCLA} working papers
in phonetics}. UCLA Los Angeles, CA 22(48). 48--76.

\leavevmode\vadjust pre{\hypertarget{ref-Lehfeldt1975}{}}%
Lehfeldt, W. 1975. {Die Verteilung der Phonemanzahl in den nat{ü}rlichen
Sprachen}. \emph{Phonetica} 31. 247--287.

\leavevmode\vadjust pre{\hypertarget{ref-vanLeijenhorst2005}{}}%
Leijenhorst, Dick C van \& Th P Van der Weide. 2005. A formal derivation
of heaps' law. \emph{Information Sciences}. Elsevier 170(2-4). 263--272.

\leavevmode\vadjust pre{\hypertarget{ref-Lindblom1990}{}}%
Lindblom, Björn. 1990. Explaining phonetic variation: A sketch of the
h\&h theory. In W. J. Hardcastle \& A. Marchal (eds.), \emph{Speech
production and speech modelling}, 403--439. Springer.
https://doi.org/\url{https://doi.org/10.1007/978-94-009-2037-8_16}.

\leavevmode\vadjust pre{\hypertarget{ref-Macklin-CordesRound2020}{}}%
Macklin-Cordes, Jayden L \& Erich R Round. 2020. Re-evaluating phoneme
frequencies. \emph{Frontiers in Psychology}. Frontiers Media SA 11.
570895.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Martindale_etal1996}{}}%
Martindale, Colin, S. M. Gusein‐Zade, Dean McKenzie \& Mark Yu.
Borodovsky. 1996. Comparison of equations describing the ranked
frequency distributions of graphemes and phonemes. \emph{Journal of
Quantitative Linguistics}. Routledge 3(2). 106--112.

@@

\leavevmode\vadjust pre{\hypertarget{ref-MielkeHume2006}{}}%
Mielke, Jeff \& Elizabeth Hume. 2006. Distinctive features. In
\emph{Encylopedia of langauge}. Elsevier.

\leavevmode\vadjust pre{\hypertarget{ref-Milivcka2009}{}}%
Milička, Jiři. 2009. Type-token \& hapax-token relation: A combinatorial
model. \emph{Glottotheory}. De Gruyter Akademie Forschung 2(1). 99--110.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Piantadosi2014}{}}%
Piantadosi, Steven T. 2014. Zipf's word frequency law in natural
language: A critical review and future directions. \emph{Psychonomic
bulletin \& review}. Springer 21(5). 1112--1130.

\leavevmode\vadjust pre{\hypertarget{ref-Potter1945}{}}%
Potter, Ralph K. 1945. Visible patterns of sound. \emph{Science}.
American Association for the Advancement of Science 102(2654). 463--470.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Snyman1975}{}}%
Snyman, J. W. 1975. \emph{{Zu\textbar'hoasi Fonologie en Woordeboek}}.
Cape Town: Balkema.

\leavevmode\vadjust pre{\hypertarget{ref-Stevens1972}{}}%
Stevens, Kenneth N. 1972. The quantal nature of speech: Evidence from
articulatory-acoustic data. In Edward E David \& Peter B Denes (eds.),
\emph{Human communication: A unified view}. McGraw-Hill.

\leavevmode\vadjust pre{\hypertarget{ref-TambovtsevMartindale2007}{}}%
Tambovtsev, Yuri \& Colin Martindale. 2007. Phoneme frequencies follow a
yule distribution. \emph{SKASE Journal of Theoretical Linguistics}.
Citeseer 4(2). 1--11.

>>>>

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran\textsuperscript{1,2} \& Nicholas A.
Lester\textsuperscript{3}\\
\textsuperscript{1}Institute of Biology, University of Neuchâtel,
Neuchâtel, Switzerland\\
\textsuperscript{2}Department of Anthropology, University of Miami,
Coral Gables, FL, USA\\
\textsuperscript{3}Department of Linguistics, University of North Texas,
Denton, TX, USA\\
OrcID, SM, 0000-0002-3969-6549\\
Corresponding authors:
\href{mailto:steven.moran@unine.ch}{\nolinkurl{steven.moran@unine.ch}};
\href{mailto:nicholas.a.lester@gmail.com}{\nolinkurl{nicholas.a.lester@gmail.com}}}
\date{04 August, 2023}

@@

How many speech sounds are there in the world's languages? To answer
this question, we extend two formal models of lexical diversity to
segmental phonology. Zipf's law states that most word types are rare;
Herdan-Heaps' law states that vocabulary size grows indefinitely with
increasingly new text. We find that both models produce accurate
estimates of the diversity of contrastive speech sounds (phonemes) in
existing typological databases. Our findings suggest that (a) most
phonemes are difficult to detect because they are rare, and (b)
increased documentation of phonological inventories always yields more
phoneme types. To check this, we apply an unrelated estimator of species
diversity from biology, Chao-Shen, and show that its estimates also
align with Herdan-Heaps' law. We use both models to extrapolate phoneme
counts to undocumented languages, resulting in an estimate of 4600-4900
distinct phonemes in the full set of languages thought to be spoken
today (\textasciitilde7-8k), which is roughly 1.5 time more than what is
currently documented in the world's languages (\textasciitilde35\%). By
extrapolating to hypothetical sample sizes, e.g., the number of
languages estimated to have been spoken through time, and beyond, we
find no indication of convergence: infinite languages mean infinite
phonemes.

@@

Assuming a target of 7151 known languages,\footnote{\url{https://www.ethnologue.com/guides/how-many-languages}?}
our models predict a range of total segment types from 4592 to 4850
(mean = 4721). Based on PHOIBLE, our current best guess of the diversity
of documented phoneme types is 3183. Therefore, the models argue for the
existence of approximately 1500 yet to be documented phoneme types. As
we probe increasingly larger and theoretically motivated language sample
sizes, i.e., languages that existed \emph{through time} (e.g., estimates
including 30-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
64-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
Herdan-Heaps' model predicted a continual increase in the number of
unique phonemes, reaching values as high as 19k. The Chao-Shen model
produced much more conservative estimates for samples of size 10k and
above, even appearing to settle on an asymptote of roughly 5500 phonemes
for samples of 100k languages or more. However, when we compare the
predictions of Herdan-Heaps and Chao-Shen against known type counts in a
closed dataset, the source of the discrepancy emerged: as the sample
grows, Chao-Shen lags behind Herdan-Heaps in approximating the true
value. Thus, given a larger pool of training data, we expect Chao-Shen
to approximate more closely the estimates of Herdan-Heaps' law.

We therefore provisionally conclude three things: (1) adding languages
to a sample will always carry a non-zero probability of producing a
novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at
which new phonemes are discovered per language will always decrease; and
(3) the decrease in the rate of new observations will never reach 0.
Thus, based on the evidence presented here, we have no reason to believe
that there is an upper bound on the number of unique phonemes possible
in spoken languages. Infinite linguistic diversity seems to bring about
infinite possible phonemes. But what would it mean if there were indeed
infinitely many potential phonemes?

If one adopts a distinctive feature-based perspective on the phonetic
targets of phonemes, then there is an \emph{a priori} upper bound of
\(2^F\) contrastive sounds, where \(F\) is the number of feature
categories and 2 reflects the choice between presence (+) or absence (-)
of that feature.\footnote{In some distinctive feature sets, some values
  may be 0, i.e., undefined, underspecified, or ``for which a segment
  can be said `not to care''' (\protect\hyperlink{ref-Hayes2009}{Hayes
  2009}, pg. 91). For example, speech sounds that are -coronal are
  underspecified for the features +/-anterior and +/-distributed, which
  capture coronal distinctions for places of articulation involving
  (inter-)dentals and alveolars vs palatal alveolars and retroflexes
  (anterior), and laminal vs apical coronals (distributed).} Clements
(\protect\hyperlink{ref-Clements2009}{2009}) coined this calculation
\emph{Feature Bounding}. The expected number of unique phonemes is
controlled by the number of features, where additional features increase
the pool of possible distinctive segments (nearly) exponentially. For
example, given a feature set such as in PHOIBLE with 37 distinctive
features, an upper bound on the number of segments under a strictly
binary formalization is over 137 billion (2\^{}37) feature vectors. In
general, distinctive feature theories of different theoretical stances
propose around two dozen features
(\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume 2006}), which
still has the potential to generate over 16 million distinct feature
vectors (2\^{}24). Crucially, this theoretical bound on phoneme
diversity does not in itself preclude the possibility of infinitely many
phonemes. To get there, one would only need to discover infinitely many
distinctive features. But, as pointed out by Lindblom
(\protect\hyperlink{ref-Lindblom1990a}{1990}), most proponents of
feature-based analysis assume a fixed, small-ish inventory of features.
The question would then be: how does the feature pool grow empirically
with the addition of new languages? Does it converge as assumed?
Unfortunately, this question presents a host of new challenges, not the
least of which concerns how to define the features themselves.

Most widely accepted feature sets define features that are specified at
different levels of phonetic detail: some features are more abstract
(i.e., general similarities such as +/- syllabic), while others are more
concrete (i.e., closer to the articulatory reality, e.g., advanced
tongue root). Some are grounded in pure articulatory or acoustic
phonetics, while others involve phonological properties of language. And
looming in the background of these discussions, the debate of whether
distinctive features are innate and universal or rather learned and
emergent continues. Thus, the precise limits on what may be included as
a feature are underspecified and a matter of some controversy (for more
fundamental critiques of features and how they are defined, see Lindblom
(\protect\hyperlink{ref-Lindblom1990a}{1990})). Current consensus and
critiques aside, there does not appear to be a principled limit to the
degree of phonetic specificity that can be encoded by a feature. Making
finer and finer contrasts may always produce more and more features.
Thus, the exponent in the Feature Bounding calculation can itself (by
hypothesis) be infinite, in line with the predictions of the models
presented here, and therefore with the notion of infinite possible
phonemes. That said, contemporary feature sets can capture phonological
classes for place and manner of articulation, which neatly classify the
phonological patterns and processes in and across languages (cf. Mielke
(\protect\hyperlink{ref-Mielke2008}{2008})).

We have so far seen that the feature space is a discrete re-coding of
the continuous variability of the physical acoustic and articulatory
spaces -- and so it should by way of finer and finer specification, be
able to support an infinitely diverse set of phonemes. However, one
objection to this account is that humans are subject to physiological
limits, both in production and perception. In speech production,
articulatory gestures are executed with some degree of imprecision
(\protect\hyperlink{ref-Ladefoged_etal1972}{Ladefoged et al. 1972};
\protect\hyperlink{ref-LindblomSundberg1971}{Lindblom \& Sundberg
1971}). If this stochastic variability is greater than the difference
between two or more phonetic targets, then no reliable distinction can
be sustained (cf. Stevens (\protect\hyperlink{ref-Stevens1972}{1972})).
In comparison, if the range of articulatory gestures is greater than
their acoustic (or perceptual) reception, there is a ``many-to-one''
mapping between articulation and perception. In perception, the ability
to discriminate acoustic stimuli is also limited both anatomically and
cognitively. Differences in the signal that fall below the various
thresholds for discrimination are thus ignored.

@@

Given that all speakers use perceptually contrastive cues (whether
signed, spoken or written), how these continuous signals map to discrete
elements is a core question of language science that goes beyond what
the phonetics-phonology interface may or may not be. Thus, we believe
the approaches presented here have broad applications to the study of
other areas of grammar and to the cross-linguistic study of languages.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BaezaNavarro2000}{}}%
Baeza--Yates, Ricardo \& Gonzalo Navarro. 2000. Block addressing indices
for approximate text retrieval. \emph{Journal of the American Society
for Information Science} 51(1). 69--82.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ChaoChiu2016}{}}%
Chao, Anne \& Chun-Huo Chiu. 2016. Species richness: Estimation and
comparison. In \emph{{Wiley StatsRef: Statistics Reference Online}},
1--26. John Wiley \& Sons.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Egghe2007}{}}%
Egghe, Leo. 2007. Untangling {Herdan's} law and {Heaps'} law:
Mathematical and informetric arguments. \emph{Journal of the American
Society for Information Science and Technology} 58(5). 702--709.

\leavevmode\vadjust pre{\hypertarget{ref-EvansLevinson2009}{}}%
Evans, Nicholas \& Stephen C. Levinson. 2009. The myth of language
universals: Language diversity and its importance for cognitive science.
\emph{Behavioral and Brain Sciences} 32(5). 429--448.

\leavevmode\vadjust pre{\hypertarget{ref-Everett2018}{}}%
Everett, Caleb. 2018. The similar rates of occurrence of consonants
across the world's languages: A quantitative analysis of phonetically
transcribed word lists. \emph{Language Sciences} 69. 125--135.

\leavevmode\vadjust pre{\hypertarget{ref-FirchowFirchow1969}{}}%
Firchow, Irwin \& Jacqueline Firchow. 1969. An abbreviated phoneme
inventory. \emph{Anthropological Linguistics} 11. 271--276.

\leavevmode\vadjust pre{\hypertarget{ref-GoodCysouw2013}{}}%
Good, Jeff \& Michael Cysouw. 2013. Languoid, doculect, and glossonym:
Formalizing the notion {``language.''} \emph{Language Documentation \&
Conservation} 7. 331--359.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Herdan1964}{}}%
Herdan, Gustav. 1964. \emph{Quantitative linguistics}. London:
Butterworth.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Joos1948}{}}%
Joos, Martin. 1948. Acoustic phonetics. \emph{Language} 24(2). 5--136.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Ladefoged_etal1972}{}}%
Ladefoged, Peter, Joseph DeClerk, Mona Lindau \& George Papcun. 1972. An
auditory-motor theory of speech production. \emph{{UCLA} working papers
in phonetics} 22(48). 48--76.

\leavevmode\vadjust pre{\hypertarget{ref-Lehfeldt1975}{}}%
Lehfeldt, Werner. 1975. {Die Verteilung der Phonemanzahl in den
nat{ü}rlichen Sprachen}. \emph{Phonetica} 31. 247--287.

\leavevmode\vadjust pre{\hypertarget{ref-vanLeijenhorst2005}{}}%
Leijenhorst, Dick C van \& Th P Van der Weide. 2005. A formal derivation
of {Heaps'} law. \emph{Information Sciences} 170(2-4). 263--272.

\leavevmode\vadjust pre{\hypertarget{ref-Lindblom1990a}{}}%
Lindblom, Björn. 1990. On the notion of {``possible speech sound.''}
\emph{Journal of Phonetics} 18(2). 135--152.

\leavevmode\vadjust pre{\hypertarget{ref-LindblomSundberg1971}{}}%
Lindblom, Björn \& Johan Sundberg. 1971. Acoustical consequences of lip,
tongue, jaw, and larynx movement. \emph{The Journal of the Acoustical
Society of America} 50(4B). 1166--1179.

\leavevmode\vadjust pre{\hypertarget{ref-Macklin-CordesRound2020}{}}%
Macklin-Cordes, Jayden L \& Erich R Round. 2020. Re-evaluating phoneme
frequencies. \emph{Frontiers in Psychology} 11. 570895.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Martindale_etal1996}{}}%
Martindale, Colin, S. M. Gusein-Zade, Dean McKenzie \& Mark Yu.
Borodovsky. 1996. Comparison of equations describing the ranked
frequency distributions of graphemes and phonemes. \emph{Journal of
Quantitative Linguistics} 3(2). 106--112.

@@

\leavevmode\vadjust pre{\hypertarget{ref-MielkeHume2006}{}}%
Mielke, Jeff \& Elizabeth Hume. 2006. Distinctive features. In
\emph{{Encylopedia of Langauge}}. Elsevier.

\leavevmode\vadjust pre{\hypertarget{ref-Milivcka2009}{}}%
Milička, Jiři. 2009. Type-token \& hapax-token relation: A combinatorial
model. \emph{Glottotheory} 2(1). 99--110.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Piantadosi2014}{}}%
Piantadosi, Steven T. 2014. Zipf's word frequency law in natural
language: A critical review and future directions. \emph{Psychonomic
Bulletin \& Review} 21(5). 1112--1130.

\leavevmode\vadjust pre{\hypertarget{ref-Potter1945}{}}%
Potter, Ralph K. 1945. Visible patterns of sound. \emph{Science}
102(2654). 463--470.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Snyman1975}{}}%
Snyman, J. W. 1975. \emph{{Zu\textbar'hoasi Fonologie en Woordeboek}}.
Cape Town: A. A. Balkema.

\leavevmode\vadjust pre{\hypertarget{ref-Stevens1972}{}}%
Stevens, Kenneth N. 1972. The quantal nature of speech: Evidence from
articulatory-acoustic data. In Edward E David \& Peter B Denes (eds.),
\emph{Human communication: A unified view}. New York: McGraw-Hill.

\leavevmode\vadjust pre{\hypertarget{ref-TambovtsevMartindale2007}{}}%
Tambovtsev, Yuri \& Colin Martindale. 2007. Phoneme frequencies follow a
{Yule} distribution. \emph{SKASE Journal of Theoretical Linguistics}
4(2). 1--11.
